---
title: "ENVS 193DS Homework 3"
author: "Jess Rousselle"
date: 2025-05-30
format:
  html:
    toc: true # use this to display a table of contents
---

<https://github.com/jarousselle/ENVS-193DS_spring-2025_final>

# Setup

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(dplyr)
library(ggplot2)
sst <- read.csv("data/SST_update2023.csv")
```

# Problem 4 Affective and exploratory visualizations

### a.Transparent statistical methods

In part 1, they used a linear regression. In part 2, they used an ANOVA test.

### b. More information needed

My co-worker should also include the results of a Tukey HSD test because this will explain which particular groups are actually different. In addition, they should also include the results of $\eta^2$ test to show the effect size, or the proportion of variation in the response that is explained by the predictor variable.

### c. Suggestions for rewriting

part 1: The annual total nitrogen load (increased/decreased) as distance from headwater increased (linear regression, F~Degrees of Freedom~ = test statistic, R^2^ = model fit, p = 0.03).

part 2: We found a (large/medium/small) difference ($\eta^2$ = effect size) between sources (urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands) in mean nitrogen load (kg year-1) (one-way ANOVA, F(degrees of freedom) = test statistic, p = 0.02, $\alpha$ = significance level). On average, (source type) tended to have (higher/lower) nitrogen load than (source type) (results from Tukey HSD: average difference in nitrogen load, 95% CI)...(more comparisons between groups can be added based off of Tukey HSD test.)

# Problem 2. Data visualization

```{r}
sst_clean <- sst |> 
  mutate(year = year(date)) |> 
 mutate(month = month(date, label = TRUE)) |> 
   group_by(month, year) |> 
   summarise(sst_clean, mean_monthly_sst = mean(temp))
  
```

